
\documentclass[usenames,dvipsnames,aspectratio=169,11pt,handout]{beamer}
%handout,
%aspectratio=43

\usepackage{others/discpreamble}
\usepackage{natbib}
\usepackage[normalem]{ulem}
\usepackage{colortbl, xcolor}
\definecolor{RoyalBlue}{rgb}{0.19, 0.55, 0.91}
\linespread{1}
\usepackage{mathrsfs}
\usepackage{csquotes}

% small bibliography
\let\oldthebibliography=\thebibliography
\renewcommand{\thebibliography}[1]{
    \oldthebibliography{#1}
    \setlength{\itemsep}{2pt}
    \tiny
}

\begin{document}

\begin{frame}[noframenumbering,plain]
	\maketitle
\end{frame}

\begin{frame}
	\frametitle{Epistemic game theory in 30 seconds}

	Game theory = individual preferences (tastes and beliefs) + equilibrium analysis.

	\vfill

	Equilibrium = rationality (expected utility) + common belief in rationality.

	\vfill

	Behavioural economics introduces “behavioural heuristics” into games.

	\vfill

	How these heuristics relate to classical preferences + equilibrium is often unclear.

	\vfill

	Epistemic game theory takes these sums \underline{\textbf{very seriously}}.
\end{frame}

\begin{frame}\frametitle{Motivation of this paper}

	Dynamic refinements (e.g. subgame perfection) use the chronological structure of games.

	\vfill

	These refinements often mispredict behaviour in classic experiments.

	\vfill

	Question: how robust are these dynamic concepts to small doubts about rationality?
\end{frame}

\begin{frame}\frametitle{Framework and concepts}

	Finite multistage games with observable actions.

	\vfill

	Players hold conjectures over others’ strategies.

	\vfill

	Dynamic rationalizability: weak (initial), backward (induction), strong (forward induction).

	\vfill

	Introduce \(p\)-belief: an event is believed with probability at least \(p<1\). \citep{mondererApproximatingCommonKnowledge1989}

	\vfill

	Define weak, backward, strong \(p\)-rationalizability.

\end{frame}

\begin{frame}\frametitle{Main results: rationalizability}

	For \(p=1\): \(p\)-versions coincide with classical notions.

	\vfill

	For any \(p<1\):
	strong and backward \(p\)-rationalizability \textbf{collapse to weak (initial) rationalizability}.

	\vfill

	Interpretation:
	even tiny doubts about higher-order rationality destroy the refinement power of dynamic reasoning.

\end{frame}

\begin{frame}\frametitle{Behavioural implications}

	With \(p<1\), dynamic refinements lose bite; predictions become essentially static.

	\vfill

	Centipede game:
	cooperation in all but the last stage is \(p\)-rationalisable.

	\vfill

	Finitely repeated Prisoner’s Dilemma:
	all threshold strategies except “cooperate in the final round” are \(p\)-rationalisable.

	\vfill

	Small doubts about rationality can rationalise non-inductive play in these games.

	\vfill

	\textbf{But I add}: for any value of \( p \) you could probably rationalise many things!

\end{frame}

\begin{frame}\frametitle{Comments}

	Results induce a conceptual dilemma:

	\vfill

	\begin{itemize}
		\item Are there better dynamic refinements we did not develop yet?
		\item Is the notion of \( p \)-belief in this paper inadequate?
		\item (\textbf{Pessimistic}): there is no hope.
	\end{itemize}

	\vfill

	How much is off-path behaviour responsible?

	\vfill

	Structural rationality \citep{siniscalchiStructuralRationalityDynamic2022} may help.

\end{frame}

\begin{frame}[noframenumbering,plain]

	\frametitle{References}

	%\nocite{*}
	\bibliography{chapters/references}
	\bibliographystyle{apacite}

\end{frame}

\end{document}