\chapter*{Preamble}\label{ch:preamble} % no number
\addcontentsline{toc}{chapter}{Preamble} % include in the ToC

These notes accompany the first part of the PhD microeconomics sequence. They cover \textbf{choice under uncertainty} and \textbf{general equilibrium theory}.
The write-up is still a work in progress, and I will continue to update it. If you spot any mistakes or typos, please let me know.

I have aimed for a conversational style rather than the more formal tone of, say, \citet{mas-colellMicroeconomicTheory1995}. I assume no prior knowledge of the topics, though—as usual—some mathematical maturity helps (and I hope you will develop it along the way!). Each lecture summarizes what we cover in class, followed by exercises and suggestions for further reading. Whenever a result is proved, I have tried to give the simplest proof available. This often makes explanations and proofs a bit longer than strictly necessary, but, I hope, also more accessible.

Before diving in, you might enjoy some non-technical background that helps frame the topics we will study: \citet[ch.~1]{krepsNotesTheoryChoice1988}, \citet[pp.~ix–xi]{debreuTheoryValueAxiomatic1959}, \citet[pp.~1–7]{myersonGameTheoryAnalysis1997}, and \citet[chs.~1–2]{gilboaTheoryDecisionUncertainty2009}.

\begin{techremark}
	You will occasionally see smaller text like this. These remarks are not essential for following the main exposition, but they add context or point to related ideas. Feel free to skip them on a first pass.
	\normalsize
\end{techremark}

These notes draw on several sources. The main reference is \citet{mas-colellMicroeconomicTheory1995}, but both here and in the text you will find pointers to alternative or complementary readings. A short reading list follows. If you would like more references or wish to discuss any of the material, just send me an email—I am always happy to talk.



\bigskip

\begin{center}
	\Large Have fun!\normalsize
\end{center}

\bigskip

\noindent
\begin{minipage}[t]{0.48\textwidth}
	\paragraph{Choice under uncertainty.}
	\begin{itemize}
		\item \textbf{\citet{mas-colellMicroeconomicTheory1995}, ch.~6.}
		\item \citet{krepsNotesTheoryChoice1988}, chs.~4-6.
		\item \citet{fishburnUtilityTheoryDecision1970}, ch.~8.
		\item \citet{krepsMicroeconomicFoundations2013}, chs.~5-6.
		\item \citet{gilboaTheoryDecisionUncertainty2009}.
	\end{itemize}
\end{minipage}%
\hfill
\begin{minipage}[t]{0.52\textwidth}
	\paragraph{General equilibrium theory.}
	\begin{itemize}
		\item \textbf{\citet{mas-colellMicroeconomicTheory1995} chs.~15–17.}
		\item \cite{thomsonFairAllocationRules2011}, sec 4.3 (no-envy).
		\item \citet{krepsMicroeconomicFoundations2013}, chs.~14–15.
		\item \citet{debreuTheoryValueAxiomatic1959}.
		\item \citet{hildenbrandIntroductionEquilibriumAnalysis1976}.
	\end{itemize}
\end{minipage}


\bibliographystyle{apacite}
\bibliography{references}
